A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2025-11-27 inst/hankin-exponential_BT.tex
typo
more work

A Plackett-Luce observation is of the form $p_{(1)}\succ
p_{(2)}\succ\cdots\succ p_{(n)}$, where $p_{(i)}$ denotes the
competitor in place $i $ for $1\leqslant i\leqslant n$.  If competitor
$j$ has strength $\beta x^{j}$, then the probability of observing
$p_1\succ p_2\succ\cdots\succ p_n$ will be, according to
equation~\ref{plackettluce},

@@

(here $[n]_x!$ is the $q$-factorial).  We see that
$f(1)=\frac{1}{n!}$, $f'(x)>0$ and $f(0)=1$; so small $x$ corresponds
to a high probability of observing the ``natural order'' in which
competitor $j$ places $j$-th, $1\leqslant j\leqslant n$.  Further, we
observe that $f(x)=1-(n-1)x + {\mathcal O}(x^2)$.  It is interesting
to note that the probability of observing the reverse order, that is
$p_n\succ p_{n-1}\succ\cdots\succ p_1$, is $f(1/x)$ for $x>0$.

@@

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{123all}  % 123all.pdf produced by exponential_BT.Rmd
\caption{Log-likelihood functions \label{123all} for support
  $\in\left\lbrace 1, 2,3\right\rbrace\times\mathbb{R}$}
\end{centering}
\end{figure}

@@

Although figure~\ref{1234third} shows a perfectly respectable
log-likelihood function, in practice we often do not know the value of
$r$, and need a joint likelihood function for $(r,x)$.  It is
convenient to normalize $r$ and use $s=r/n$ for consistency between
different values of $n$, and to treat $s$ as a continuous variable.
The only reasonable way to calulate such a likelihood function appears
to be numerical; fortunately this is straightforward.

We leverage the {\tt hyper2} R package~\citep{hankin2017_rmd} which
includes extensive numerical functionality to simulate Plackett-Luce
situations, and in this case function {\tt rrace()} simulates a race
between competitors of known Plackett-Luce strengths.  To obtain a
likelihood function for a specific value of $(x,s)$ at a particular
observed rank $r$, we run repeated {\em in-silico} races between
competitors of Plackett-Luce strengths $x^1,x^2,\ldots x^n$, and
observe the frequency with which competitor $\lceil ns\rceil$ places
$r$.  For example, consider $x=0.1, r=3$.  We might simulate this as
follows:

\label{rsession}
\begin{verbatim}
n <- 9  # number of competitors
s <- 4  # observation: focal competitor places fourth

@@

The maximum likelihood estimate for $s$ would therefore be $4/9\simeq
0.44$, conditional on $x=0.1$.  To estimate $x$ we would need repeated
observations, and here we note that small values of $x$ lead to
underdispersed results, for the focal competitor would come $r$-th
with probability $\geqslant x^1+x^2+\ldots x^r$

To validate the suggested likelihood method, we set $x=0.85$ and
$r=3$, that is, competitor $i$ has strength $\propto 0.85^i,
1\leqslant i\leqslant 9$; the focal competitor has the third strongest
Plackett-Luce strength of $0.85^3\simeq 0.61$.  The results of running
{\tt rrace()} 100 times with $n=9$ are given in
table~\ref{results100insilico}.

@@

Now, we can run repeated simulations like those on page~\pageref{rsession}, but
with $n=10^4$ trials; as support we use $x$ running from 0.1 to 1 and
$r=1,2,\ldots 9$: figure~\ref{toyexample}.  We see reasonable
agreement between the true values of $x$ and $r$ and the estimates.
Note the odd shape of the contours and the non-independence of the
parameters.

@@

Formula 1 racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  In a typical season, $\sim 20$
competitors race at $\sim 20$ venues and here we consider the 2023
season, using Lewis Hamilton as the focal competitor.  His results are
shown in table~\ref{lewishamilton}; Qatar and the US are disregarded
as he did not finish.

@@

Again using numerical simulation, amounting to repeated evaluation of
generalised version of equations \ref{aftersim} and \ref{requalfour}
(this time with $b=10^4$ trials per point), we obtain
figure~\ref{formula1}.  From this, we see that $r$ is less than about
0.25 and $x$ is between about 0.4 and 0.8.  Recall that, in this
model, $x=0.5$ implies that the Plackett-Luce strength of each
competitor is about twice that of the next.

@@

Here a simplification of the Plackett-Luce likelihood functions for
order statistics was proposed, some simple properties obtained, and
inference problems from three disciplines given.  The results seem to
be difficult to obtain any other way.  The assumptions furnish a
reasonable, intuitive, and computationally tractable likelihood
function with wide applicability.  Further work might include the
analysis of more than one focal competitor.

>>>>

A Plackett-Luce observation is of the form $p_{(1)}\succ
p_{(2)}\succ\cdots\succ p_{(n)}$, where $p_{(i)}$ denotes the
competitor in place $i$ for $1\leqslant i\leqslant n$.  If competitor
$j$ has strength $\beta x^{j}$, then the probability of observing
$p_1\succ p_2\succ\cdots\succ p_n$ for fixed $n$ will be, according to
equation~\ref{plackettluce},

@@

(here $[n]_x!$ is the $q$-factorial).  We see that $f(0)=1$, $f'(x)<0$
and $f(1)=\frac{1}{n!}$; so small $x$ corresponds to a high
probability of observing the ``natural order'' in which competitor $j$
places $j$-th, $1\leqslant j\leqslant n$.  Further, we observe that
$f(x)=1-(n-1)x + {\mathcal O}(x^2)$.  If $x=1/2$, it can be shown that
the probability of observing the natural order is asymptotically $\sim
2^{-n}$.  It is interesting to note that the probability of observing
the reverse order, that is $p_n\succ p_{n-1}\succ\cdots\succ p_1$, is
$f(1/x)$ for $x>0$.

The rest of this paper considers the following inference problem: We
consider $n$ competitors with Bradley-Terry strengths proportional to
$x^i, 1\leqslant i\leqslant n$.  Our object of inference is the focal
competitor with strength $\propto x^r$, with $r$ unknown.  The
competitors are ranked randomly according to a Plackett-Luce model; we
observe only the rank of the focal competitor.  What may we infer
about $(x,r)$?

@@

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{123all}  % 123all.pdf produced by exponential_BT.Rmd
\caption{The case $n=3$.  Three log-likelihood
  functions \label{123all}, one for each of the three possible
  observations of the rank $r$ of the focal competitor, presented as a
  function of $\left\lbrace 1, 2,3\right\rbrace\times\mathbb{R}^-$.
  (a), $r=1$; (b), $r=2$; (c), $r=3$}
\end{centering}
\end{figure}

@@

Although figure~\ref{1234third} shows a perfectly respectable
log-likelihood function, in practice we often do not know the value of
$r$, and need a joint likelihood function for $(r,x)$.  It is
convenient to normalize $r$ and use $s=r/n$ for consistency between
different values of $n$, and to treat $s$ as a continuous variable.
The only reasonable way to calculate such a likelihood function
appears to be numerical; fortunately this is straightforward.

We leverage the {\tt hyper2} R package~\citep{hankin2017_rmd} which
includes extensive numerical functionality to simulate Plackett-Luce
situations, and in this case function {\tt rrace()} simulates a race
between competitors of known Plackett-Luce strengths.  To obtain a
likelihood function for a specific value of $(x,s)$ at a particular
observed rank $r$, we run repeated {\em in-silico} races between
competitors of Plackett-Luce strengths $x^1,x^2,\ldots x^n$, and
observe the frequency with which competitor $\lceil ns\rceil$ places
$j$.  For example, consider $x=0.1, j=4$: each competitor has 10 times
the strength of the next, and the focal competitor placed fourth.  We
might simulate this as follows:

\label{rsession}
\begin{verbatim}
n <- 9 # number of competitors
       # observation: focal competitor places fourth

@@

The maximum likelihood estimate for $s$ would therefore be $4/9\simeq
0.44$, conditional on $x=0.1$.  To estimate $x$ we would need repeated
observations, and here we note that small values of $x$ lead to
underdispersed results.

To validate the suggested likelihood method, we set $x=0.85$ and
$r=3$, that is, competitor $i$ has strength $\propto 0.85^i,
1\leqslant i\leqslant 9$; the focal competitor has $r=3$, that is, has
the third strongest Plackett-Luce strength of $0.85^3\simeq 0.61$.
The results of running {\tt rrace()} 100 times with $n=9$ are given in
table~\ref{results100insilico}.

@@

Now, we can run repeated simulations like those on
page~\pageref{rsession}, but with $n=10^4$ trials; as support we use
$x$ running from 0.1 to 1 and $r=1,2,\ldots 9$.
Figure~\ref{toyexample} gives a visual representation of the
log-likelihood.  We see reasonable agreement between the true values
of $x$ and $r$, indicated with a red dot, and the estimates.  Note the
odd shape of the contours and the non-independence of the parameters.

@@

Formula 1 racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  In a typical season, ${\sim} 20$
competitors race at ${\sim} 20$ venues and here we consider the 2023
season, using Lewis Hamilton as the focal competitor.  His results are
shown in table~\ref{lewishamilton}; Qatar and the US are disregarded
as he did not finish.

@@

Again using numerical simulation, amounting to repeated evaluation of
generalised version of equations \ref{aftersim} and \ref{requalfour}
(this time with $b=10^4$ trials per point), we obtain
figure~\ref{formula1}.  From this, we see that $s$ is less than about
0.25 and $x$ is between about 0.4 and 0.8.  Recall that, in this
model, $x=0.5$ implies that the Plackett-Luce strength of each
competitor is about twice that of the next.

@@

Here a parametrized Plackett-Luce likelihood function for order
statistics was proposed, some simple properties obtained, and
inference problems from motor sport given.  The results seem to
be difficult to obtain any other way.  The assumptions furnish a
reasonable, intuitive, and computationally tractable likelihood
function with wide applicability.  Further work might include the
analysis of more than one focal competitor.
